Beating Simple Backtracking Search in Subgraph Algorithms\thanks{This work was supported by the Engineering and Physical Sciences Research Council [grant numbers EP/026842/1 and EP/M508056/1].}

### 2018-11-29 paper/paper.tex
Edit Ruth's edits

Although the problem is \NP-complete, by combining design techniques from artificial intelligence
with careful algorithm engineering, modern subgraph isomorphism solvers can often produce exact
solutions quickly even on graphs with thousands of vertices. The current single strongest subgraph
isomorphism solver uses ``highest degree first'' as a value-ordering heuristic to direct a
constraint programming style search
\cite{DBLP:conf/cp/McCreeshP15,DBLP:conf/lion/KotthoffMS16,DBLP:journals/jair/McCreeshPST18}. This
heuristic is much better than branching randomly, but is still far from perfect. To offset mistakes
made by this heuristic, this paper proposes a new perspective on value-ordering: rather than
defining a search order, we use degree to direct what \emph{proportion} of the search effort should
be spent in each subproblem. By combining rapid restarts and nogood recording, and introducing a
small amount of randomness into the value-ordering heuristic, we make a state-of-the-art subgraph
algorithm perform two orders of magnitude better on a large number of satisfiable instances, whilst
performing worse rarely on satisfiable instances, and never on unsatisfiable instances.  This
strategy is also effective in an optimisation setting, producing benefits in two maximum common
induced subgraph algorithms.

This new form of search can also be parallelised, with a much simpler implementation than
conventional work-stealing. By running many threads with different random seeds but the same restart
schedule, and sharing nogoods only following restarts, we can achieve speedups of thirty one from a
thirty six core machine, or ninety five by using ten such machines with message passing.

@@

Solution-biased search allows for a much simpler parallel implementation. We
create a number of threads, and give each thread its own random seed, otherwise each thread
performs the same sequential search. Then, whenever a thread restarts, it uses a barrier to wait for
every other thread to also restart. At each restart each thread's new nogood can be gathered, and
then each thread resumes, with a larger set of nogood than in a sequential run. Finally, whenever
any single thread terminates, either due to having found a solution or proved unsatisfiability,
then every other thread may immediately terminate.

@@

It is important to emphasise that there is no expectation of a linear speedup from this approach,
particularly for satisfiable instances. Due to the biased random nature of the value-ordering heuristic,
one of our extra threads may ``get lucky'' and find a solution much quicker than we would
sequentially, leading to a superlinear speedup. Conversely, our extra threads may contribute nothing to finding a
solution---or worse, due to new nogoods altering the choice made by the biased random branching heuristic,
we could even get an absolute slowdown. We can see both of these effects in the bottom left plot of
\cref{figure:parallel}: although we achieve roughly a linear speedup on unsatisfiable instances,
satisfiable instances show much greater variability. However, this approach does at least mirror our intuition of
allocating search effort in proportion to where the value-ordering heuristic believes it will be
most fruitful, and so we should not be too surprised that we see roughly a linear speedup on average
on harder instances.

>>>>

Although the problem is \NP-complete, by combining design techniques from artificial intelligence
with careful algorithm engineering, modern subgraph isomorphism solvers can often produce exact
solutions quickly even on graphs with thousands of vertices. The current single strongest subgraph
isomorphism solver uses ``highest degree first'' as a value-ordering heuristic to direct a
constraint programming style search
\cite{DBLP:conf/cp/McCreeshP15,DBLP:conf/lion/KotthoffMS16,DBLP:journals/jair/McCreeshPST18}. This
heuristic is much better than branching randomly, but is still far from perfect. To offset mistakes
made by this heuristic, this paper proposes a new perspective on value-ordering: rather than
defining a search order, we use degree to direct what \emph{proportion} of the search effort should
be spent in each subproblem. By combining rapid restarts and nogood recording, and introducing a
small amount of randomness into the value-ordering heuristic, we make a state-of-the-art subgraph
algorithm perform two orders of magnitude better on a large number of satisfiable instances, whilst
performing worse only rarely on satisfiable instances, and never on unsatisfiable instances.  This
strategy is also effective in an optimisation setting, producing benefits in two maximum common
induced subgraph algorithms.

This new form of search can also be parallelised, with a \emph{much} simpler implementation than
conventional work-stealing. By running many threads with different random seeds but the same restart
schedule, and sharing nogoods only following restarts, we can achieve speedups of thirty one from a
thirty six core machine, or ninety five by using ten such machines with message passing.

@@

Solution-biased search allows for a much simpler parallel implementation. We
create a number of threads, and give each thread its own random seed, otherwise each thread
performs the same sequential search. Then, whenever a thread restarts, it uses a barrier to wait for
every other thread to also restart. At each restart each thread's new nogoods can be gathered, and
then each thread resumes, with a larger set of nogoods than in a sequential run. Finally, whenever
any single thread terminates, either due to having found a solution or proved unsatisfiability,
then every other thread may immediately terminate.

@@

It is important to emphasise that there is no expectation of a linear speedup from this approach,
particularly for satisfiable instances. Due to the biased random nature of the value-ordering heuristic,
one of our extra threads may ``get lucky'' and find a solution much quicker than we would
sequentially, leading to a superlinear speedup. Conversely, our extra threads may contribute nothing to finding a
solution---or worse, due to new nogoods altering the choice made by the random branching heuristic,
we could even get an absolute slowdown. We can see both of these effects in the bottom left plot of
\cref{figure:parallel}: although we achieve roughly a linear speedup on unsatisfiable instances,
satisfiable instances show much greater variability. However, this approach does at least mirror our intuition of
allocating search effort in proportion to where the value-ordering heuristic believes it will be
most fruitful, and so we should not be too surprised that we see roughly a linear speedup on average
on harder instances.
